$Z_3$ topological order in the quantum dimer-pentamer model

### 2016-12-06 qdpm_paper.tex
going to pull

We have introduced the quantum dimer-pentamer model on the square lattice and studied the ground state at the RK point. Using numerical calculations, we have demonstrated that the RK state of the QDPM is a dimer liquid, without spontaneously broken symmetry. The extensive topological degeneracy of the square lattice QDM has been reduced to a finite 9-fold degeneracy in the QDPM on a torus. We find that the dimer correlations decay exponentially with distance in contrast to the power-law decay of dimer correlations in the QDM. Additionally we also have demonstrated that monomer defects are deconfined in the QDPM, in contrast with the critical confinement of monomers in the QDM. Finally, we find that the $Z_3$ magnetic string correlations decay exponentially and display no evidence of $Z_3$ vortex condensation; this is consistent with $Z_3$ vortices being gapped quasiparticles excitations above the QDPM ground state. These results suggest a the low energy physics of the QDPM is described by a $Z_3$ gauge theory with a $Z_3$ topological ordered ground state. Thus we have demonstrated that systems with Ising degrees of freedom and local constraints can lead to $Z_N$ topological order for $N>2$, providing the a new stable phase of matter beyond the commonly seen $Z_2$ topological spin liqud.

@@

Finally, given the stability of the topologically ordered phase, it may be possible to realize the QDPM in a more microscopically realistic spin model with two body interactions. The local constraints of the QDM can be realized in two-body spin-$1/2$ and Bose-Hubbard models, and the QDM Hamiltonian can arise perturbatively in the constrained low energy Hilbert space~\cite{Balents2002a,Zhitomirsky2005,Isakov2006c,Albuquerque2008}. Thus future work, it may be possible to engineer the QDPM in the low energy subspace of an experimentally realizable two-body model, providing a route to the experimental realization of a $Z_3$ topologically ordered spin liquid.

@@

Computations were performed on the Vermont Advanced Computing Core supported by NASA (NNX-08AO96G).

@@

To test for the possibility of translational and rotational symmetry breaking corresponding to the features in $S_d$ at the Brillouin zone edges, we define several order parameters which would capture such potential symmetry breaking. For rotational symmetry breaking we define 
\begin{equation}
\Mrot \equiv  \frac{2}{L^2} \left(N_h - N_v\right) \label{eq:Mrot},
\end{equation}
where $N_v$ ($N_h$) is the total number of vertical (horizontal) dimers; note that $\exval{M_{\rm{rot}}^2 } \neq 0$ implies there is rotational symmetry breaking. to To capture translational symmetry breaking by one lattice site in a direction perpendicular to the majority of the dimers we define $\Mperp$ 
\begin{align}
\Mperp \equiv  \frac{8}{L^2} \sum_{\Box} \Bigl(  &(-1)^{r_x} \theta\left(\Mrot \right) \Pi_\Box^v \notag\\
+ &(-1)^{r_y}\theta\left(-\Mrot \right) \Pi_\Box^h\Bigr) \label{eq:Mperp},
\end{align}
where $\theta(x)$ is the Heaviside function, and $\Pi_\Box^h$ ($\Pi_\Box^v$) is one if the plaquette $\Box$ has two horizontal (vertical) dimers, and zero otherwise. Similarly, we define $\Mpar$ to be given by \Eqref{eq:Mperp} with $r_x$ and $r_y$ swapped between terms. Additionally, we define order parameters that indicate ``columnar" and ``staggered" dimer order~\cite{Syljuasen2005}:
\begin{align}
\MstagSq \equiv  \frac{1}{L^2} \biggl[ &\biggl( \sum_{\ell_h} (-1)^{r_x+r_y} n_d \left(\vec{r}_{\ell_h}\right) \biggr)^2  \notag\\
+&\biggl( \sum_{\ell_v} (-1)^{r_x+r_y} n_d \left(\vec{r}_{\ell_v}\right) \biggr)^2\biggr]\label{eq:Mstag},
\end{align}
\begin{align}
\McolSq \equiv  \frac{1}{L^2} \biggl[  &\biggl( \sum_{\ell_h} (-1)^{r_x} n_d \left(\vec{r}_{\ell_h}\right) \biggr)^2 \notag \\
+ &\biggl( \sum_{\ell_v} (-1)^{r_y} n_d \left(\vec{r}_{\ell_v}\right) \biggr)^2  \biggr]\label{eq:Mcol},
\end{align}
where the sums over $\ell_h$ ($\ell_v$) are taken over all horizontal (vertical) links. In \figref{fig:order_params} we have plotted the finite-size scaling of these order parameters; we find in all cases that while finite size systems display symmetry breaking due to finite-size effects, these order parameters all vanish in the thermodynamic limit, as demonstrated by the extrapolation in \figref{fig:order_params}.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\linewidth]{order_params_plot.pdf}
    \caption{The symmetry breaking order parameters.}
    \label{fig:order_params}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

>>>>

We have introduced the quantum dimer-pentamer model on the square lattice and studied the ground state at the RK point. Using numerical calculations, we have demonstrated that the RK state of the QDPM is a dimer liquid, without spontaneously broken symmetry. The extensive topological degeneracy of the square lattice QDM has been reduced to a finite 9-fold degeneracy in the QDPM on a torus. We find that the dimer correlations decay exponentially with distance in contrast to the power-law decay of dimer correlations in the QDM. Additionally we also have demonstrated that monomer defects are deconfined in the QDPM, in contrast with the critical confinement of monomers in the QDM. Finally, we find that the $Z_3$ magnetic string correlations decay exponentially and display no evidence of $Z_3$ vortex condensation; this is consistent with $Z_3$ vortices being gapped quasiparticles excitations above the QDPM ground state.

The above results suggest a the low energy physics of the QDPM is described by a $Z_3$ gauge theory with a $Z_3$ topological ordered ground state. Thus we have demonstrated that systems with Ising degrees of freedom and local constraints can lead to $Z_N$ topological order for $N>2$, providing the a new stable phase of matter beyond the commonly seen $Z_2$ topological spin liqud.

@@

Finally, given the stability of the topologically ordered phase, it may be possible to realize the QDPM in a more microscopically realistic spin model with two body interactions. The local constraints of the QDM can be realized in two-body spin-$1/2$ and Bose-Hubbard models, and the QDM Hamiltonian can arise perturbatively in the constrained low energy Hilbert space~\cite{Balents2002a,Zhitomirsky2005,Isakov2006c,Albuquerque2008}. Thus future work, it may be possible to engineer the QDPM in the low energy subspace of an experimentally realizable two-body model.

@@

Our calculations were performed on the VACC....

@@

To test for the possibility of translational and rotational symmetry breaking corresponding to the features in $S_d$ at the Brillouin zone edges, we define several order parameters which would capture such potential symmetry breaking. For rotational symmetry breaking we define 
\begin{equation}
\Mrot \equiv  \frac{2}{L^2} \left(N_v - N_h\right) \label{eq:Mrot},
\end{equation}
where $N_v$ ($N_h$) is the total number of vertical (horizontal) dimers; note that $\exval{M_{\rm{rot}}^2 } \neq 0$ implies there is rotational symmetry breaking. to To capture translational symmetry breaking by one lattice site in a direction perpendicular to the majority of the dimers we define $\Mperp$ 
\begin{align}
\Mperp \equiv  \frac{8}{L^2} \sum_{\Box} \Bigl(  &(-1)^{r_x} \theta\left(\Mrot \right) \Pi_\Box^v \notag\\
+ &(-1)^{r_y}\theta\left(-\Mrot \right) \Pi_\Box^h\Bigr) \label{eq:Mperp},
\end{align}
where $\theta(x)$ is the Heaviside function, and $\Pi_\Box^h$ ($\Pi_\Box^v$) is one if the plaquette $\Box$ has two horizontal (vertical) dimers, and zero otherwise. Similarly, we define $\Mpar$ to be given by \Eqref{eq:Mperp} with $r_x$ and $r_y$ swapped between terms. Additionally, we define order parameters that indicate ``columnar" and ``staggered" dimer order~\cite{Syljuasen2005}:
\begin{align}
\MstagSq \equiv  \frac{4}{L^4} \biggl[ &\biggl( \sum_{\ell_h} (-1)^{r_x+r_y} n_d \left(\vec{r}_{\ell_h}\right) \biggr)^2  \notag\\
+&\biggl( \sum_{\ell_v} (-1)^{r_x+r_y} n_d \left(\vec{r}_{\ell_v}\right) \biggr)^2\biggr]\label{eq:Mstag},
\end{align}
\begin{align}
\McolSq \equiv  \frac{4}{L^4} \biggl[  &\biggl( \sum_{\ell_h} (-1)^{r_x} n_d \left(\vec{r}_{\ell_h}\right) \biggr)^2 \notag \\
+ &\biggl( \sum_{\ell_v} (-1)^{r_y} n_d \left(\vec{r}_{\ell_v}\right) \biggr)^2  \biggr]\label{eq:Mcol},
\end{align}
where the sums over $\ell_h$ ($\ell_v$) are taken over all horizontal (vertical) links. In \figref{fig:order_params} we have plotted the finite-size scaling of these order parameters; we find in all cases that while finite size systems display symmetry breaking due to finite-size effects, these order parameters all vanish in the thermodynamic limit, as demonstrated by the extrapolation in \figref{fig:order_params}.
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[]
    \centering
    \includegraphics[width=1.0\linewidth]{order_params_plot.pdf}
    \caption{The symmetry breaking order parameters.}
    \label{fig:order_params}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
